---
title: "scrnaboxRMD"
author: "Saeid Amiri"
date: "2022-08-09"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages({
    library(scrnaboxR)
    library(Seurat)
    library(dplyr)
    library(cowplot)
    library(ggplot2)
    library(pheatmap)
    library(enrichR)
    library(rafalib)
    library(msigdbr)
    library(tidyr)
    library(tidyverse)
    library(fgsea)
    library(clusterProfiler)
    library(enrichplot)
    library(org.Hs.eg.db)
    AnnotationDbi::keytypes(org.Hs.eg.db)
    library(Seurat)
    library(clusterProfiler)
})

```

# content
-   Cell marker genes
-   Working with DGE Contrast 
   -  Volcano plot
   -  split contrast
   -  Calculate distance between contrast
   -  ditt plot



## Cell marker genes
Download `/gdp` from step7 on your working path `WDP`, cell marker genes and seurat object are  the `de_genes` and  `seu.int`, respectively


```{r}
WDP='/Volumes/F/for_my_website/scrnabox0/related_to_scrnaR'
setwd(WDP)
mk_genes<-readRDS(paste0(WDP, "/gdp/de_genes.rds"))
seu.int<-readRDS(paste0(WDP, "/gdp/seu.int.c.rds"))
```


Select top 2
```{r}
mk_genes %>%
    group_by(cluster) %>%
    top_n(-2, p_val_adj) -> top_sel
```

```{r}
top_sel
```
See the plot of genes: 
```{r}
mypar(3, 3, mar = c(5, 5, 3, 3))
for (i in unique(top_sel$cluster)) {
    barplot(sort(setNames(top_sel$avg_log2FC, top_sel$gene)[top_sel$cluster == i], F),
        horiz = T, las = 1, main = paste0(i, " vs. rest"),)
    abline(v = c(0, 0.25))
}
```
Visualize via heatmap 

```{r}
seu.int_sel <- ScaleData(seu.int, features = as.character(unique(top_sel$gene)), assay = "RNA")
DoHeatmap(seu.int_sel, features = as.character(unique(top_sel$gene)), group.by = "cell.types.pool",    assay = "RNA")
```   

Visualize the overall expression via dot plot. 

```{r}
 DotPlot(seu.int_sel, features = rev(as.character(unique(top_sel$gene))), group.by = "cell.types.pool",
    assay = "RNA") + coord_flip()
```

Visulaize via violin plot 
 
```{r}
VlnPlot(seu.int_sel, features = as.character(unique(top_sel$gene)), ncol = 5, group.by = "cell.types.pool", assay = "RNA")
```

## Working with DGE Contrast 
Download the contrasts relust from step 7, 
### Volcano plot
```{r}
library(EnhancedVolcano)
cont_a53t<-read.csv(paste0(WDP,'/cont_main/design1.csv'))
EnhancedVolcano(cont_a53t,lab = rownames(cont_a53t),x = 'logFC',  y = 'P.Value', FCcutoff=1, title = 'A53T - rest', drawConnectors = TRUE, max.overlaps= 30)
ggsave(file = "EnhancedVolcano2.pdf")
```

```{r}
library(htmlwidgets)
library(plotly)
cont_a53t_sig<-cont_a53t[cont_a53t$adj.P.Val<0.05,]
p <- plot_ly(data = cont_a53t_sig, x = cont_a53t_sig$logFC, y = -log(cont_a53t_sig$adj.P.Val), text = cont_a53t_sig$X, mode = "markers") %>% 
  layout(title ="A53T vs rest")
saveWidget(p, "A53T_vs_rest.html")
```


### Spliting
If you want to split the contrast into up and down, you can run the following codes
```{r}
file_path=paste0(WDP,'/cont_inte')
dest_path=paste0(WDP,'/cont_split')
scrnaboxR::split_up_down(file_path,dest_path,cut_pvalue=0.05,cut_log_low=-0.25,cut_log_up=0.25, n_row=200)
```

### Distance between contrasts
```{r}
file0<-extract_file(dest_path)
con<-file0[[2]]
```

```{r echo=FALSE, results=FALSE}
dst<-dist_contrast(con)
dim <- ncol(dst)
```

```{r}
#png(file = paste0(WDP,"/heatmap.png"), bg = "transparent")
image(1:dim, 1:dim, dst, axes = FALSE, xlab="", ylab="")
label0<-file0[[1]]
axis(1, 1:dim,label0 , cex.axis = 0.5, las=3)
axis(2, 1:dim, label0, cex.axis = 0.5, las=1)
text(expand.grid(1:dim, 1:dim), sprintf("%0.1f", dst), cex=0.6)
#dev.off()
```

### Dit plot
```{r}
gwas_genes<-read.csv(file=paste0(WDP,'/all_PD_gwas_loci_genes_filtered.csv'))$hgnc_symbol
file_path=paste0(WDP,'/cont_split')
des_path=paste0(WDP,'/Output')
dittplot(gwas_genes,file_path,des_path)
```

## Gene Set Enrichment Analysis (GSEA)


```{r}
genes_rank <- setNames(mk_genes$avg_log2FC, casefold(rownames(mk_genes),upper = T))
head(genes_rank)
msigdbgmt <- msigdbr::msigdbr("Homo sapiens")
msigdbgmt <- as.data.frame(msigdbgmt)
```
You check the  list  of available gene sets
```{r}
unique(msigdbgmt$gs_subcat)
```
Select the subset that you need.
```{r}
msigdbgmt_subset <- msigdbgmt[msigdbgmt$gs_subcat == "GO:BP", ]
gmt <- lapply(unique(msigdbgmt_subset$gs_name), function(x) {
    msigdbgmt_subset[msigdbgmt_subset$gs_name == x, "gene_symbol"]
})
names(gmt) <- unique(paste0(msigdbgmt_subset$gs_name, "_", msigdbgmt_subset$gs_exact_source))
gmt %>% head() %>% lapply(head)
```


conduct enrichemnt analysis and order according the ES
```{r}
fgseaRes <- fgsea(pathways = gmt, stats = genes_rank, minSize = 15, maxSize = 500, scoreType = "pos")
fgseaRes <- fgseaRes[order(fgseaRes$ES, decreasing = T), ]
```

```{r}
barplot(sort(genes_rank, decreasing = T))
head(fgseaRes[order( -abs(NES)), ], n=10)
```

get enrichment score plot for a specific pathway
```{r}
plotEnrichment(gmt[["GOBP_ACTIN_FILAMENT_BASED_PROCESS_GO:0030029"]], genes_rank)
```

You can plot enrichment for multiple pathways.
```{r}
top_up <- fgseaRes %>% filter(ES > 0) %>% top_n(10, wt=-padj)
top_down <- fgseaRes %>% filter(ES < 0) %>% top_n(10, wt=-padj)
top_pathways <- bind_rows(top_up, top_down) %>% arrange(-ES)
plotGseaTable(gmt[top_pathways$pathway], genes_rank, fgseaRes, gseaParam = 0.5)
```

The following barplot can be used to compare Hallmark pathways (obtained from  https://www.biostars.org/p/467197/)

```{r}
fgseaResTidy <- fgseaRes %>% as_tibble() %>% arrange(desc(NES))
fgseaResTidy %>% dplyr::select(-leadingEdge, -ES) %>% arrange(padj) %>% DT::datatable()
#ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) + geom_col(aes(fill=padj<0.05)) +coord_flip() +labs(x="Pathway", y="Normalized Enrichment Score",title="Hallmark pathways NES from GSEA") + theme_minimal()
```



## Gene Ontology gene sets (KEGG gene sets)


```{r}
setwd(paste0(WDP,'/Output'))
file0='design1down'
sel_genes <- read.csv(paste0(WDP,'/cont_split/',file0,'.csv',sep=""))[,1]
sel_genes_go <- clusterProfiler::enrichGO(sel_genes,"org.Hs.eg.db",keyType = "SYMBOL",ont = "BP",minGSSize = 50)
write.csv(sel_genes_go@result,file = paste(file0,'enrichGO','.csv',sep=''))
enr_go <- clusterProfiler::simplify(sel_genes_go)
enrichplot::emapplot(enrichplot::pairwise_termsim(enr_go),showCategory = 30, cex_label_category = 0.5)
ggsave(file = "enrich1.pdf")
dotplot(enrichplot::pairwise_termsim(enr_go))
ggsave(file = "dot1.pdf")
cnetplot(enrichplot::pairwise_termsim(enr_go)) 
ggsave(file = "cnet1.pdf")

gmt <- msigdbr::msigdbr(species = "human", category = "H")
sel_go0 <- clusterProfiler::enricher(gene = sel_genes,,pAdjustMethod = "BH",pvalueCutoff  = 0.05, qvalueCutoff  = 0.05, TERM2GENE = gmt[,c("gs_name", "gene_symbol")])
write.csv(sel_go0@result,file = paste(file0,'enricher','.csv',sep=''))
```


The follwoing run the KEGG pathway enrichment analysis. 
```{r}
sel_genes_id<-mapIds(org.Hs.eg.db, sel_genes, 'ENTREZID', 'SYMBOL',multiVals=last)
get_ENTREZID <- function(x){
  ab<-NULL
  for (i in 1: length(x)){
    ab[i]<-x[[i]]
  }
  message("There are ", sum(is.na(ab)), " NA and they are dropped")
  ret<-ab
  }
genes_id<-get_ENTREZID(sel_genes_id)
enrich_kegg <- enrichKEGG(gene = genes_id)
head(enrich_kegg, n=10)
```

